---
title: "Russian blues do not reveal effects of language on color discrimination"
author: "Sihan Chen, Bevil Conway, Edward Gibson"
date: '2023-02-21'
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tidyverse)
library(brms)
library(ggpubr)
set.seed(20230220)
```
# Load data

We used the original data exclusion criteria.
```{r}
df <- read.csv('russian_blues_data_processed_modified_2.csv')
df_ok <- df %>% filter(include_by_performance == 1,
                                         include_by_subject == 1,
                                         include_by_stim == 1) %>% mutate(left = as.numeric(left),
                                                                          right = as.numeric(right)) %>% 
  rowwise() %>% 
  mutate(pair = paste0(min(left, right), '_', max(left, right))) %>% ungroup()

summary <- df_ok %>% group_by(language, distance, condition, category, subject) %>%
  summarise(mean.RT = mean(response_time),
            std.RT = sd(response_time))
```

In this document, we repeat the analyses in Winawer et al. (2007), but instead of using their original two-way categorization method, where stimuli pairs were simply categorized by whether they crossed a participant's color boundary or not, we use our proposed three-way categorization method, where stimuli pairs were categorized by whether 1) they were both categorized as light blues (within light), 2) they were both categorized as dark blues (within dark), or 3) one was categorized as dark blue, and another as light blue (between).

# Repeat analysis

## Analysis with all the data
> For each group, there was a highly significant main effect of distance: in Russian speakers [926 vs. 1,245 msec, near color vs. far color; F > (1, 20) = 267; P < 0.001] and English speakers [800 vs. 1,078 msec; F (1, 20) = 144.1; P < 0.001]. 
>
> `r tufte::quote_footer('--- Winawer et al. (2007)')`

```{r}
ggplot(summary, aes(x=distance, y = mean.RT)) +
  facet_wrap(~language) +
  stat_summary(geom = 'col', position = 'dodge', fun = 'mean', fill='white', color='black') +
  stat_summary(geom = 'errorbar', position = 'dodge',
              fun.data = 'mean_cl_boot', alpha = 0.40)

effect_of_distance_english <- brm(response_time ~ distance + (1 + distance | subject) + (1 | pair), 
                                  data = df_ok %>% filter(language == 'english'),
                                  family = gaussian(),
                                  warmup = 2000,
                                  iter = 10000,
                                 chain = 4,
                                 thin = 10)
summary(effect_of_distance_english)

effect_of_distance_russian <- brm(response_time ~ distance + (1 + distance | subject) + (1 | pair), 
                                  data = df_ok %>% filter(language == 'russian'),
                                  family = gaussian(),
                                  warmup = 2000,
                                  iter = 10000,
                                 chain = 4,
                                 thin = 10)
summary(effect_of_distance_russian)
```

> Russian speakers were slower overall than English speakers.

```{r}
ggplot(summary, aes(x=language, y = mean.RT)) +
  stat_summary(geom = 'col', position = 'dodge', fun = 'mean', fill='white', color='black') +
  stat_summary(geom = 'errorbar', position = 'dodge',
              fun.data = 'mean_cl_boot', alpha = 0.40)

effect_of_language <- brm(response_time ~ language + (1 | subject) + (1 + language | pair), 
                                  data = df_ok,
                                  family = gaussian(),
                                  warmup = 2000,
                                  iter = 10000,
                                 chain = 4,
                                 thin = 10)
summary(effect_of_language)
```

## Analysis on only Russian speakers
> there was a category advantage under both the no- interference and the spatial-interference conditions, but not under the verbal 
> interference condition

```{r}
ggplot(summary %>%
         mutate(category = ifelse(category=='within_lightblue', 'within light', category),
                category = ifelse(category=='within_darkblue', 'within dark', category),
           category = factor(category, levels = c("within light", "between", "within dark")),
           language = ifelse(language=='english', 'English', 'Russian')),
       aes(x = category, y = mean.RT, fill = category)) +
  facet_wrap(~condition) +
  stat_summary(geom = 'col', position = 'dodge', fun = 'mean') +
   stat_summary(geom = 'errorbar', position = 'dodge',
              fun.data = 'mean_cl_boot', alpha = 0.40) +
  coord_cartesian(ylim=c(700,1150)) +
  theme_bw(25) +
  theme(legend.key.size = unit(2, 'cm'),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        strip.background = element_rect(fill='white', color = 'gray')) +
  scale_fill_manual(values = c('#89CFF0','gray50','#00008B')) 


#TODO run these analyses. Idea: verbal interference is the baseline here  
category_interference_russian_1 <- brm(response_time ~ category * condition + (1 + category * condition | subject) +
                                       (1 + category * condition | pair),
                                     data = df_ok %>% filter(language == 'russian',
                                                             category %in% c("between", "within_lightblue"),
                                                             condition %in% c('no_interference', 'verbal_interference')),
                                      family = gaussian(),
                                      warmup = 2000,
                                      iter = 10000,
                                     chain = 4,
                                     thin = 10)
category_interference_russian_1 %>% summary()

category_interference_russian_2 <- brm(response_time ~ category * condition + (1 + category * condition | subject) +
                                       (1 + category * condition | pair),
                                     data = df_ok %>% filter(language == 'russian',
                                                             category %in% c("between", "within_darkblue"),
                                                             condition %in% c('no_interference', 'verbal_interference')),
                                      family = gaussian(),
                                      warmup = 2000,
                                      iter = 10000,
                                     chain = 4,
                                     thin = 10)
category_interference_russian_2 %>% summary()

category_interference_russian_3 <- brm(response_time ~ category * condition + (1 + category * condition | subject) +
                                       (1 + category * condition | pair),
                                     data = df_ok %>% filter(language == 'russian',
                                                             category %in% c("between", "within_lightblue"),
                                                             condition %in% c('spatial_interference', 'verbal_interference')),
                                      family = gaussian(),
                                      warmup = 2000,
                                      iter = 10000,
                                     chain = 4,
                                     thin = 10)
category_interference_russian_3 %>% summary()

category_interference_russian_4 <- brm(response_time ~ category * condition + (1 + category * condition | subject) +
                                       (1 + category * condition | pair),
                                     data = df_ok %>% filter(language == 'russian',
                                                             category %in% c("between", "within_darkblue"),
                                                             condition %in% c('spatial_interference', 'verbal_interference')),
                                      family = gaussian(),
                                      warmup = 2000,
                                      iter = 10000,
                                     chain = 4,
                                     thin = 10)
category_interference_russian_4 %>% summary()

```

> This effect was completely due to the near-color condition (Fig. 3), supported by a significant three-way interaction among category, interference, and distance

```{r}
ggplot(summary %>%
         mutate(category = ifelse(category=='within_lightblue', 'within light', category),
                category = ifelse(category=='within_darkblue', 'within dark', category),
           category = factor(category, levels = c("within light", "between", "within dark")),
           language = ifelse(language=='english', 'English', 'Russian')),
       aes(x = category, y = mean.RT, fill = category)) +
  facet_grid(cols = vars(condition), rows = vars(distance)) +
  stat_summary(geom = 'col', position = 'dodge', fun = 'mean') +
   stat_summary(geom = 'errorbar', position = 'dodge',
              fun.data = 'mean_cl_boot', alpha = 0.40) +
   coord_cartesian(ylim=c(500,1400)) +
  theme_bw(25) +
  theme(legend.key.size = unit(2, 'cm'),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        strip.background = element_rect(fill='white', color = 'gray')) +
  scale_fill_manual(values = c('#89CFF0','gray50','#00008B')) 


category_interference_distance_russian_1 <-  brm(response_time ~ category * condition * distance + (1 + category * condition * distance | subject) +
                                               (1 + category * condition | pair),
                                              data = df_ok %>% filter(language == 'russian',
                                                                     category %in% c("between", "within_lightblue"),
                                                                     condition %in% c('no_interference', 'verbal_interference')),
                                              family = gaussian(),
                                              warmup = 2000,
                                              iter = 10000,
                                              chain = 4,
                                              thin = 10)

category_interference_distance_russian_1 %>% summary()

category_interference_distance_russian_2 <-  brm(response_time ~ category * condition * distance + (1 + category * condition * distance | subject) +
                                               (1 + category * condition | pair),
                                              data = df_ok %>% filter(language == 'russian',
                                                                     category %in% c("between", "within_darkblue"),
                                                                     condition %in% c('no_interference', 'verbal_interference')),
                                              family = gaussian(),
                                              warmup = 2000,
                                              iter = 10000,
                                              chain = 4,
                                              thin = 10)

category_interference_distance_russian_2 %>% summary()

category_interference_distance_russian_3 <-  brm(response_time ~ category * condition * distance + (1 + category * condition * distance | subject) +
                                               (1 + category * condition | pair),
                                              data = df_ok %>% filter(language == 'russian',
                                                                     category %in% c("between", "within_lightblue"),
                                                                     condition %in% c('spatial_interference', 'verbal_interference')),
                                              family = gaussian(),
                                              warmup = 2000,
                                              iter = 10000,
                                              chain = 4,
                                              thin = 10)

category_interference_distance_russian_3 %>% summary()

category_interference_distance_russian_4 <-  brm(response_time ~ category * condition * distance + (1 + category * condition * distance | subject) +
                                               (1 + category * condition | pair),
                                              data = df_ok %>% filter(language == 'russian',
                                                                     category %in% c("between", "within_darkblue"),
                                                                     condition %in% c('spatial_interference', 'verbal_interference')),
                                              family = gaussian(),
                                              warmup = 2000,
                                              iter = 10000,
                                              chain = 4,
                                              thin = 10)

category_interference_distance_russian_4 %>% summary()
```

> In near-color trials, Russian speakers showed a category advantage without interference and with spatial interference

```{r}
# {within light blue, between}
category_no_interference_near_1 <- brm(response_time ~ category + 
                                       (1 + category | subject) +
                                       (1 + category  | pair),
                                              data = df_ok %>% 
                                       filter(language == 'russian',
                                              category %in% c("between", "within_lightblue"),
                                              condition %in% c('no_interference'),
                                              distance == 'near'),
                                     family = gaussian(),
                                     warmup = 2000,
                                     iter = 10000,
                                     chain = 4,
                                     thin = 10)

category_no_interference_near_1 %>% summary()

# {within dark blue, between}
category_no_interference_near_2 <- brm(response_time ~ category + 
                                       (1 + category | subject) +
                                       (1 + category  | pair),
                                              data = df_ok %>% 
                                       filter(language == 'russian',
                                              category %in% c("between", "within_darkblue"),
                                              condition %in% c('no_interference'),
                                              distance == 'near'),
                                     family = gaussian(),
                                     warmup = 2000,
                                     iter = 10000,
                                     chain = 4,
                                     thin = 10)


category_no_interference_near_2 %>% summary()

# {within light blue, between}
category_spatial_interference_near_1 <- brm(response_time ~ category + 
                                              (1 + category | subject) + 
                                              (1 + category | pair),
                                             data = df_ok %>%
                                               filter(language=='russian',
                                                      category %in% c("between", "within_lightblue"),
                                                      condition %in% c('spatial_interference'),
                                                      distance == 'near'),
                                             family = gaussian(),
                                             warmup = 2000,
                                             iter = 10000,
                                             chain = 4,
                                             thin = 10)
category_spatial_interference_near_1 %>% summary()

# {within dark blue, between}
category_spatial_interference_near_2 <- brm(response_time ~ category + 
                                              (1 + category | subject) + 
                                              (1 + category | pair),
                                             data = df_ok %>%
                                               filter(language=='russian',
                                                      category %in% c("between", "within_darkblue"),
                                                      condition %in% c('spatial_interference'),
                                                      distance == 'near'),
                                             family = gaussian(),
                                             warmup = 2000,
                                             iter = 10000,
                                             chain = 4,
                                             thin = 10)
category_spatial_interference_near_2 %>% summary()
```

> In near-color trials, the difference in the category advantage between no interference and verbal interference was significantly greater for Russian than English speakers.

Here we define two types of category advantage:
CA1: category advantage = RT in Within Light category - RT in Between category -, for each participant
CA2: category advantage = RT in Within Dark category - RT in Between category, for each pariticipant
```{r}
# Calculate category advantage
ca <- summary %>% filter(distance == 'near') %>% 
  select(-std.RT) %>% 
  pivot_wider(names_from = category, values_from = mean.RT) %>% 
  mutate(ca1 = within_lightblue - between,
         ca2 = within_darkblue - between) 


ggplot(ca %>% pivot_longer(cols = c(ca1, ca2), names_to = 'ca_category', values_to = 'value'), aes(x = language, y = value)) +
  facet_grid(cols=vars(condition), rows=vars(ca_category))+
  stat_summary(fun = 'mean', geom = 'col') +
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar', alpha = 0.5, width = 0.4) +
  theme_bw(25) +
  theme(legend.key.size = unit(2, 'cm'),
        panel.grid = element_blank(),
        panel.border = element_blank(),
        strip.background = element_rect(fill='white', color = 'gray'))


# Calculate difference in CA1 
ca1_diff <- ca %>% select(language, distance, condition, subject, ca1) %>%
  pivot_wider(names_from = condition, values_from = ca1) %>% 
  mutate(diff_no_int_ver_int = no_interference - verbal_interference,
         diff_spa_int_ver_int = spatial_interference - verbal_interference) %>%
  select(language, distance, subject, diff_no_int_ver_int, diff_spa_int_ver_int)

ggplot(ca1_diff, aes(x = language, y = diff_no_int_ver_int)) +
  stat_summary(fun = 'mean', geom = 'col') +
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar', alpha = 0.5, width = 0.4) +
  stat_compare_means(comparisons = list(c("english", "russian")), method = 't.test') +
  theme_bw(25) +
  ylab('Difference in CA1 between No Interference and Verbal Interference (ms)') +
  xlab('') +
  theme(panel.grid = element_blank())

ggplot(ca1_diff, aes(x = language, y = diff_spa_int_ver_int)) +
  stat_summary(fun = 'mean', geom = 'col') +
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar', alpha = 0.5, width = 0.4) +
  stat_compare_means(comparisons = list(c("english", "russian")), method = 't.test') +
  theme_bw(25) +
  ylab('Difference in CA1 between No Interference and Verbal Interference (ms)') +
  xlab('') +
  theme(panel.grid = element_blank())

# Calculate difference in CA2
ca2_diff <- ca %>% select(language, distance, condition, subject, ca2) %>%
  pivot_wider(names_from = condition, values_from = ca2) %>% 
  mutate(diff_no_int_ver_int = no_interference - verbal_interference,
         diff_spa_int_ver_int = spatial_interference - verbal_interference) %>%
  select(language, distance, subject, diff_no_int_ver_int, diff_spa_int_ver_int)

ggplot(ca2_diff, aes(x = language, y = diff_no_int_ver_int)) +
  stat_summary(fun = 'mean', geom = 'col') +
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar', alpha = 0.5, width = 0.4) +
  stat_compare_means(comparisons = list(c("english", "russian")), method = 't.test') +
  theme_bw(25) +
  ylab('Difference in CA1 between No Interference and Verbal Interference (ms)') +
  xlab('') +
  theme(panel.grid = element_blank())

ggplot(ca2_diff, aes(x = language, y = diff_spa_int_ver_int)) +
  stat_summary(fun = 'mean', geom = 'col') +
  stat_summary(fun.data = 'mean_cl_boot', geom = 'errorbar', alpha = 0.5, width = 0.4) +
  stat_compare_means(comparisons = list(c("english", "russian")), method = 't.test') +
  theme_bw(25) +
  ylab('Difference in CA1 between No Interference and Verbal Interference (ms)') +
  xlab('') +
  theme(panel.grid = element_blank())

```

